Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-08-16 manuscript/manuscript.Rmd
Tidying up some more

A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds.
One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
Another likely cause is that machine learning models commonly overfit the training data [@mehta_high-bias_2019].
Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized [@dar_farewell_2021]. 
With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@mehta_high-bias_2019].
One frequently used method to reduce overfitting is K-fold cross validation [@raschka_model_2020], but this approach cannot be effectively employed when there is one or few time series in the training set.
To remedy the instability problem, we use an ensemble-based method, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
We found this simple ensemble technique to be an effective way to improve generalizability in the limited data regime.

@@

Overall forecasting skill scores for each model across all three scenarios are summarized in Fig 2.
Average scores (black lines) hide wide variation in forecast skill.
Generally, ML performance tends to be bracketed between MCMC (essentially the theoretical optimum), and the statistical ARIMA model, though sometimes performing worse than ARIMA or better than MCMC.
Under scenarios with alternative stable states (saddle and stochastic), the distribution of scores is often bimodal for ML models, though not MCMC. 
The ML ensemble model often performs as well as the best ML model on average.
Note that a wide prediction of uncertainty does not mean a wide range in the score skill -- for instance the ensemble model which has the widest array of outcomes often has a relatively tight distribution of score, especially in logs skill.
This reflects the relative contributions of accuracy and uncertainty as components in the forecasts.
Most ML scores are comparable to MCMC skill except for the scenario of the saddle-node bifurcation, where all other models are much worse.
Th get a deeper understanding of these general patterns, we now turn to examine each of the forecast distributions themselves in comparison to the future ensemble produced by the true generative process model, Fig 3-5.

@@

Forecasts of the Hopf bifurcation (Fig 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
However, like the other ML models, it fails to predict the growing amplitude of those oscillations.
Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
Despite this, overall all methods score comparably in CRPS score (Fig 2), with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast,
while MCMC scores are worst in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.

>>>>

A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@madhyastha_model_2019].
We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained the same model with different initialization seeds.
One explanation for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@madhyastha_model_2019]. 
Another likely cause is that machine learning models commonly overfit the training data [@mehta_high-bias_2019].
Across deep learning, overfitting is a fundamental issue, arising from neural networks being highly overparameterized [@dar_farewell_2021]. 
With so many parameters, deep learning models tend to have high variance and thus overfit the training data, a consequence of the bias-variance trade-off common across statistics and machine learning [@mehta_high-bias_2019].
One frequently used method to reduce overfitting is K-fold cross validation [@raschka_model_2020], but this approach cannot be effectively employed when there is one or few time series in the training set.
To remedy the instability problem, we use an ensemble-based method, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
We found this simple ensemble technique to be an effective way to improve generalizability in the limited data regime.

@@

Overall forecasting skill scores for each model across all three scenarios are summarized in Fig 2.
Average scores (black lines) hide wide variation in forecast skill.
Generally, ML performance tends to be bracketed between MCMC (essentially the theoretical optimum), and the statistical ARIMA model, though sometimes performing worse than ARIMA or better than MCMC.
Under scenarios with alternative stable states (saddle and stochastic), the distribution of scores is often bimodal for ML models, though not MCMC. 
The ML ensemble model often performs as well as the best ML model on average.
Note that a wide prediction of uncertainty does not mean a wide range in the score skill -- for instance the ensemble model which has the widest array of outcomes often has a relatively tight distribution of score, especially in logs skill.
This reflects the relative contributions of accuracy and uncertainty as components in the forecasts.
Most ML scores are comparable to MCMC skill except for the scenario of the saddle-node bifurcation, where all other models are much worse.
To get a deeper understanding of these general patterns, we now turn to examine each of the forecast distributions themselves in comparison to the future ensemble produced by the true generative process model, Fig 3-5.

@@

Forecasts of the Hopf bifurcation (Fig 3) are roughly comparable across the phenomenological models (ARIMA and machine learning models).
All models are trained using 100 time points drawn from the period of time prior to the onset of the Hopf bifurcation, which leads to a stable limit cycle that gradually grows in magnitude.
Most models predict a roughly constant mean with a spread roughly equal to that created by the stochastic oscillations around the stable node as seen in the training data prior to the bifurcation.
Notably, the GRU method picks up the oscillatory nature of the dynamics, despite the fact that no true oscillations were yet present in the training data.
However, like the other ML models, it fails to predict the growing amplitude of those oscillations.
Having access to the true model structure, the MCMC model alone predicts the transition into a pattern of oscillations which grows over time, though it tends to overestimate the amplitude of those oscillations initially.
Despite this, overall all methods score comparably in CRPS score (Fig 2) with most ML methods actually out-performing the MCMC score on average (Fig 5), albeit with much greater variation in individual scores.
A clearer picture can be seen by looking at these skill scores over time (Fig 6-7), which show that MCMC is initially performing worse (over-predicting variance) but as oscillations grow further, it starts outperforming the more stationary forecasts of the ML models.

@@

Examining patterns in the scores over time (Fig 6-7) provides a more nuanced understanding of the forecast dynamics than aggregate scores alone (Fig 2).
In the Hopf bifurcation, CRPS scores get worse over time across all methods, including the MCMC forecasts.
In the saddle node bifurcation and stochastic transition, the same pattern holds somewhat more dramatically for non-MCMC forecast, while MCMC scores are worse in mid-range.
Comparing CRPS scores to logs score also emphasizes the relative role of uncertainty: for instance, the MCMC scores for the Hopf bifurcation get steadily worse under CRPS but not under logs score. 
A relatively sharp transition can be seen under both MCMC scores on the Hopf bifurcation once the magnitude of the oscillations exceeds the variance created by mere stochasticity: the MCMC model no longer over-estimates the spread of the data, while the ML models now underestimate that variation. 
CRPS scores for stochastic transitions exhibit a distinct two-branch pattern, with scores for a given replicate being either very high (poor skill) or very low, reflecting whether the individual 'true' replicate matches the mean state predicted by the forecast or the other state. 
Logs skill score may be a better measure in this context, where correctly capturing the uncertainty in the forecast means that this bi-modal structure in scores can be avoided entirely, e.g. by the MCMC predictions. 
The forecast-skill-over time plots illustrate different reasons for the bi-modal distribution in skill seen for the saddle-node and stochastic transition scenarios in Fig 2 respectively: in the case of the saddle node, the two modes are distinguished by time-horizon: short term forecasts are relatively accurate, and longer term forecasts (i.e. after the catastrophic transition) are poor.
In the case of the stochastic transition model, the two modes are not structured by horizon but by replicate, with some replicates having transitioned and others still in the original state.
